Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-06-24 kappa.tex
Begin to document SIA and SSA (todos left).

% My commands
%\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
%\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
%\newcommand{\aref}[0]{\textbf{[ref.]}}
\renewcommand{\div}[1]{\mathrm{div}\,#1}
\renewcommand{\grad}[1]{\vec{\mathrm{grad}}\,#1}
\newcommand{\tensdiv}[1]{\vec{\mathrm{div}}\,#1}
\newcommand{\tensgrad}[1]{\bm{\mathrm{grad}}\,#1}

@@

Complete stress balance

\begin{equation}
    \left\{\begin{aligned}
        \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
            &= \pdv{p}{x} \\
        \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
            &= \pdv{p}{y} \\
        \pdv{\tau_{zx}}{x} + \pdv{\tau_{zy}}{y} + \pdv{\tau_{zz}}{z}
            &= \pdv{p}{z} - \rho g
    \end{aligned}\right.
\end{equation}

Shallow ice approximation

\begin{equation}
    \left\{\begin{aligned}
        \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \\
        \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \\
        0 &= \pdv{p}{z} - \rho g
    \end{aligned}\right.
\end{equation}

\begin{equation}
    \pdv{\vec{v}}{z} = 2A\,(\rho g)^n\,(s-z)^n\,|\grad{s}|^{n-1}\,\grad{s}
\end{equation}

Shallow shelf approximation

\begin{equation}
    \left\{\begin{aligned}
        \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
            &= \pdv{p}{x} \\
        \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
            &= \pdv{p}{y} \\
        \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g
    \end{aligned}\right.
\end{equation}

\begin{equation}
    \left\{\begin{aligned}
        \pdv{x} \left[2\bar{\nu}H
                      \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
            + \pdv{y} \left[\bar{\nu}H
                            \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
        \pdv{x} \left[\bar{\nu}H
                      \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
            + \pdv{y} \left[2\bar{\nu}H
                            \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
    \end{aligned}\right.
\end{equation}

\begin{equation}
    \bar{\nu} = \frac{\bar{B}}{2}\,\dot{\epsilon}^{\frac{1-n}{n}}
              = \frac{1}{2 \bar{A}^{\frac{1}{n}}}\,
                \dot{\epsilon}^{\frac{1-n}{n}}
\end{equation}

Basal sliding

>>>>

% My commands
%\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
%\newcommand{\aref}[0]{\textbf{[ref.]}}
\renewcommand{\div}[1]{\mathrm{div}\,#1}
\renewcommand{\grad}[1]{\vec{\mathrm{grad}}\,#1}
\newcommand{\tensdiv}[1]{\vec{\mathrm{div}}\,#1}
\newcommand{\tensgrad}[1]{\bm{\mathrm{grad}}\,#1}

@@

PISM, the ice sheet model that we use, employ two well-documented
approximations of ice flow equations: the shallow ice approximation and the
the shallow shelf approximation. Although both approximations can be derived
from a rigorous scaling approach, we don't expand these lengthy derivations
here, assuming instead the simplifications and focusing on their effect.

% Fig. SIA + SSA

Distinguishing the hydrostatic and deviatoric components of the stress tensor,
the balance of stresses (\eqn{stressbalance}) can be expanded into

\begin{equation}
    \label{eqn:stressbalance}
    \tensdiv{\bm\tau} - \grad{p} + \rho\,\vec{g} = \vec{0}.
\end{equation}

Projecting along the three Cartesian coordinate axis, this reads

\begin{equation}
    \label{eqn:fullstokes}
    \left\{\begin{aligned}
        \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
            &= \pdv{p}{x} \\
        \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
            &= \pdv{p}{y} \\
        \pdv{\tau_{zx}}{x} + \pdv{\tau_{zy}}{y} + \pdv{\tau_{zz}}{z}
            &= \pdv{p}{z} - \rho g
    \end{aligned}\right.
\end{equation}

The shallow ice approximation (SIA) assumes that basal friction is high enough
for the vertical shear stresses $\{\tau_{xz}, \tau_{yz}\}$ to predominate over
all other components of the deviatoric stress tensor. In this framework, the
projected stress-balance can be simplified to

\begin{equation}
    \left\{\begin{aligned}
        \pdv{\tau_{xz}}{z} &= \pdv{p}{x} \\
        \pdv{\tau_{yz}}{z} &= \pdv{p}{y} \\
        0 &= \pdv{p}{z} - \rho g
    \end{aligned}\right..
\end{equation}

This yields

\todo{develop intermediary steps}

\begin{equation}
    \pdv{\vec{v}_{SIA}}{z} = 2A\,(\rho g)^n\,(s-z)^n\,
                             |\grad{s}|^{n-1}\,\grad{s}.
\end{equation}

The shallow shelf approximation (SSA), on the opposite of the SIA, assumes that
basal friction is low enough for ice to deform predominantly by horizontal
expansion within the entire ice column. In this context, the stress-balance can
be simplified to

\begin{equation}
    \left\{\begin{aligned}
        \pdv{\tau_{xx}}{x} + \pdv{\tau_{xy}}{y} + \pdv{\tau_{xz}}{z}
            &= \pdv{p}{x} \\
        \pdv{\tau_{yx}}{x} + \pdv{\tau_{yy}}{y} + \pdv{\tau_{yz}}{z}
            &= \pdv{p}{y} \\
        \pdv{\tau_{zz}}{z} &= \pdv{p}{z} - \rho g
    \end{aligned}\right.,
\end{equation}

and horizontal velocities can be assumed independent of depth, so that

\begin{equation}
    \left\{\begin{aligned}
        \pdv{v_x}{z} &= 0 \\
        \pdv{v_y}{z} &= 0 \\
    \end{aligned}\right..
\end{equation}

This yields

\todo{develop intermediary steps}

\begin{equation}
    \left\{\begin{aligned}
        \pdv{x} \left[2\bar{\nu}H
                      \left(2\pdv{v_x}{x} + \pdv{v_y}{y}\right) \right]
            + \pdv{y} \left[\bar{\nu}H
                            \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
        \pdv{x} \left[\bar{\nu}H
                      \left(\pdv{v_x}{y} + \pdv{v_y}{x}\right) \right]
            + \pdv{y} \left[2\bar{\nu}H
                            \left(\pdv{v_x}{x} + 2\pdv{v_y}{y}\right) \right]
            + \tau_{bx} &= \rho_i gH \pdv{h}{x} \\
    \end{aligned}\right.,
\end{equation}

where $\bar{\nu}$ is the depth-integrated apparent viscosity defined by

\begin{equation}
    \bar{\nu} = \frac{1}{H}\int_b^s\nu.
\end{equation}

In PISM, the SIA and SSA velocities are added,

\begin{equation}
    \label{eqn:siassa}
    \vec{v} = \vec{v}_{SIA} + \vec{v}_{SSA}.
\end{equation}

Because $\vec{v}_{SSA}$ is small where $\vec{v}_{SIA}$ are large and
reciprocally, we can expect the model to perform well in regions where
conditions for applicability of either the SIA or the SSA are met. However,
equation~\ref{eqn:siassa} is a heuristic, and its validity in the zone of
transition between shear flow and longitudinal flow has not yet been
mathematically proven.

\subsection{Basal sliding}
